---
title: "Predicting Low Muscle Mass in U.S. Adults (NHANES 2011–2018)"
subtitle: "CIND860 Capstone Project — Toronto Metropolitan University"
author: "Adiel Mallik"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6,
  fig.align  = "center",
  cache   = FALSE
)

# ── Package installation helper ───────────────────────────────────────────────
# Run once if any package is missing; knit will find them afterwards.
required_pkgs <- c(
  "readxl", "dplyr", "tidyr", "ggplot2", "forcats",
  "mice", "caret", "glmnet", "randomForest", "xgboost",
  "kernlab", "pROC", "MLmetrics", "yardstick",
  "shapviz", "kernelshap",
  "knitr", "kableExtra", "gt",
  "DescTools", "scales"
)
new_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

# ── Libraries ─────────────────────────────────────────────────────────────────
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(mice)
library(caret)
library(glmnet)
library(randomForest)
library(xgboost)
library(kernlab)
library(pROC)
library(MLmetrics)
library(yardstick)
library(shapviz)
library(kernelshap)
library(knitr)
library(kableExtra)
library(gt)
library(DescTools)
library(scales)

set.seed(42)
```

---

# Setup and Data Loading

This capstone project applies four machine-learning classifiers to predict
**low muscle mass** — defined as an appendicular skeletal muscle mass index
(ASMI) below the sex-specific 20th percentile — in U.S. adults aged 20–59
using the NHANES 2011–2018 survey cycles.  The cleaned dataset was produced by
the companion R data-cleaning script and exported as an Excel workbook.

```{r load-data}
# ── Load data ─────────────────────────────────────────────────────────────────
DATA_PATH <- "nhanes_clean_2011_2018.xlsx"   # adjust path if needed

nhanes <- read_excel(DATA_PATH) %>%
  mutate(
    # Ensure correct types
    low_muscle_mass = as.integer(low_muscle_mass),
    low_grip        = as.integer(low_grip),
    sex             = factor(sex, levels = c("Male", "Female")),
    race_ethnicity  = factor(race_ethnicity),
    survey_cycle    = factor(survey_cycle)
  )

cat("Dataset dimensions:", nrow(nhanes), "rows ×", ncol(nhanes), "columns\n")
```

```{r head-data}
# ── First 6 rows ──────────────────────────────────────────────────────────────
kable(head(nhanes), caption = "First 6 rows of the cleaned NHANES dataset",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

```{r outcome-summary}
# ── Outcome prevalence ────────────────────────────────────────────────────────
nhanes_cc <- nhanes %>% filter(!is.na(low_muscle_mass))   # complete-case for outcome

prev_overall <- mean(nhanes_cc$low_muscle_mass) * 100
prev_sex     <- nhanes_cc %>%
  group_by(sex) %>%
  summarise(n = n(),
            n_lmm = sum(low_muscle_mass),
            pct   = round(mean(low_muscle_mass) * 100, 1),
            .groups = "drop")

cat(sprintf("Overall low-muscle-mass prevalence: %.1f %%\n", prev_overall))
print(prev_sex)
```

The dataset contains **`r nrow(nhanes)`** adults aged 20–59 drawn from four
NHANES cycles (2011–2012, 2013–2014, 2015–2016, 2017–2018).  The outcome
variable `low_muscle_mass` is a binary indicator (1 = low, 0 = normal) derived
from sex-specific 20th-percentile ASMI thresholds computed in the cleaning
script.

---

# Multiple Imputation

Several nutritional and socioeconomic variables carry moderate rates of
missingness (up to ~40 % for some dietary items).  We address this with
**Multiple Imputation by Chained Equations (MICE)** using `m = 10` imputed
datasets, pooled via Rubin's rules.  The primary outcome variables (`asmi`,
`alm_kg`, `low_muscle_mass`) are **excluded** from imputation because the
outcome must be derived from complete DXA data only.

```{r imputation-setup}
# ── Variables to impute ───────────────────────────────────────────────────────
IMPUTE_VARS <- c("poverty_ratio", "protein_g", "energy_kcal",
                 "calcium_mg", "vitamin_d_mcg")

# Variables available as predictors in the imputation model
# (exclude outcome-derived columns to avoid data leakage)
PREDICTOR_VARS <- c("age_years", "sex", "race_ethnicity", "bmi",
                    "weight_kg", "height_cm", "waist_cm", "max_grip_kg",
                    "survey_cycle",
                    "poverty_ratio", "protein_g", "energy_kcal",
                    "calcium_mg", "vitamin_d_mcg")

# Subset used for imputation — work only with predictor + impute variables
nhanes_imp_input <- nhanes %>%
  select(participant_id, all_of(PREDICTOR_VARS)) %>%
  mutate(sex           = factor(sex),
         race_ethnicity = factor(race_ethnicity),
         survey_cycle  = factor(survey_cycle))

# ── Missing data summary before imputation ────────────────────────────────────
miss_tbl <- data.frame(
  variable    = IMPUTE_VARS,
  n_missing   = sapply(IMPUTE_VARS, function(v) sum(is.na(nhanes_imp_input[[v]]))),
  pct_missing = sapply(IMPUTE_VARS, function(v) round(mean(is.na(nhanes_imp_input[[v]])) * 100, 1)),
  row.names   = NULL
)

kable(miss_tbl,
      col.names = c("Variable", "N Missing", "% Missing"),
      caption   = "Missing data in variables selected for imputation") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

```{r run-mice, cache=TRUE}
# ── Build method vector ───────────────────────────────────────────────────────
# All five imputation targets are continuous → "pmm" (predictive mean matching)
# Binary variables (none here) would use "logreg"
method_vec <- make.method(nhanes_imp_input %>% select(-participant_id))

# Override for our target variables
method_vec[IMPUTE_VARS] <- "pmm"

# Predictor matrix: suppress participant_id from being used as a predictor
pred_mat <- make.predictorMatrix(nhanes_imp_input %>% select(-participant_id))

# ── Run MICE ──────────────────────────────────────────────────────────────────
set.seed(42)
mice_obj <- mice(
  nhanes_imp_input %>% select(-participant_id),
  m        = 10,
  method   = method_vec,
  predictorMatrix = pred_mat,
  maxit    = 10,
  printFlag = FALSE,
  seed     = 42
)

cat("Imputation complete. Number of imputed datasets:", mice_obj$m, "\n")
```

```{r mice-diagnostics, fig.cap="Convergence trace plots for imputed variables (mean and SD across iterations)"}
# ── Convergence check ─────────────────────────────────────────────────────────
plot(mice_obj, layout = c(2, length(IMPUTE_VARS)),
     main = "MICE Convergence: Mean (left) and SD (right)")
```

```{r pool-imputations}
# ── Pool imputed datasets: create one analysis-ready data frame ───────────────
# Strategy: average the imputed values across m = 10 datasets (Rubin's rules
# for continuous variables), then rejoin with complete-case outcome columns.

imp_list <- lapply(seq_len(mice_obj$m), function(i) complete(mice_obj, i))

# Compute column-wise means across imputed datasets for the five imputed vars
pooled_mat <- Reduce("+", lapply(imp_list, function(d) d[, IMPUTE_VARS])) /
              mice_obj$m

nhanes_pooled <- nhanes_imp_input %>%
  select(-all_of(IMPUTE_VARS)) %>%
  bind_cols(pooled_mat)

# Rejoin DXA-derived and outcome columns (complete cases only for those)
nhanes_pooled <- nhanes_pooled %>%
  left_join(nhanes %>% select(participant_id, asmi, alm_kg,
                              low_muscle_mass, low_grip,
                              bmi_outlier, protein_outlier, energy_outlier),
            by = "participant_id")

cat("Pooled dataset dimensions:", nrow(nhanes_pooled), "×", ncol(nhanes_pooled), "\n")
cat("Remaining NAs in imputed vars:\n")
print(colSums(is.na(nhanes_pooled[, IMPUTE_VARS])))
```

The five variables were imputed using **predictive mean matching (PMM)**, which
preserves the empirical distribution of each variable and prevents implausible
values (e.g., negative nutrient intakes).  Convergence was verified visually
via trace plots above; mixing across chains indicates stable estimates.

---

# Feature Engineering

Three derived features are constructed prior to modelling:

| Feature | Description |
|---------|-------------|
| `bri` | Body Roundness Index — a proxy for central adiposity |
| `age_group` | 4-level ordinal factor (decade bands) |
| `protein_g` (winsorized) | Sex-specific 99.5th-percentile cap to reduce leverage of extreme values |

## Body Roundness Index

The BRI is computed from the Thomas et al. (2013) formula:

$$\text{BRI} = 364.2 - 365.5 \times \sqrt{1 - \left(\frac{\text{waist}_{m}}{2\pi}\right)^2 / \left(\frac{\text{height}_{m}}{2}\right)^2}$$

where waist circumference and height are both in metres.

```{r feature-engineering}
# ── 3a. Body Roundness Index (Thomas et al., 2013) ────────────────────────────
nhanes_fe <- nhanes_pooled %>%
  mutate(
    waist_m  = waist_cm  / 100,
    height_m = height_cm / 100,
    # Clamp the argument to [0,1] to guard against floating-point edge cases
    bri_arg  = pmin(pmax(1 - ((waist_m / (2 * pi))^2) / ((height_m / 2)^2), 0), 1),
    bri      = 364.2 - 365.5 * sqrt(bri_arg)
  ) %>%
  select(-waist_m, -height_m, -bri_arg)

cat("BRI summary:\n")
print(summary(nhanes_fe$bri))

# ── 3b. Age group (4-level factor) ───────────────────────────────────────────
nhanes_fe <- nhanes_fe %>%
  mutate(age_group = cut(age_years,
                         breaks = c(20, 30, 40, 50, 60),
                         labels = c("20-29", "30-39", "40-49", "50-59"),
                         right  = FALSE,
                         include.lowest = TRUE))

cat("\nAge group distribution:\n")
print(table(nhanes_fe$age_group, useNA = "ifany"))

# ── 3c. Winsorise protein_g at sex-specific 99.5th percentile ─────────────────
p995_male   <- quantile(nhanes_fe$protein_g[nhanes_fe$sex == "Male"],
                        0.995, na.rm = TRUE)
p995_female <- quantile(nhanes_fe$protein_g[nhanes_fe$sex == "Female"],
                        0.995, na.rm = TRUE)

nhanes_fe <- nhanes_fe %>%
  mutate(protein_g = case_when(
    sex == "Male"   & protein_g > p995_male   ~ p995_male,
    sex == "Female" & protein_g > p995_female ~ p995_female,
    TRUE ~ protein_g
  ))

cat(sprintf("\nProtein winsorisation caps — Male: %.1f g/day  |  Female: %.1f g/day\n",
            p995_male, p995_female))
cat("Protein range after winsorisation:", range(nhanes_fe$protein_g, na.rm = TRUE), "\n")
```

```{r bri-distribution, fig.cap="Distribution of Body Roundness Index by sex"}
ggplot(nhanes_fe %>% filter(!is.na(bri), !is.na(sex)),
       aes(x = bri, fill = sex, colour = sex)) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  scale_fill_manual(values   = c("Male" = "#3B82F6", "Female" = "#EF4444")) +
  scale_colour_manual(values = c("Male" = "#3B82F6", "Female" = "#EF4444")) +
  labs(title = "Body Roundness Index Distribution by Sex",
       subtitle = "Thomas et al. (2013) formula applied to NHANES waist and height measurements",
       x = "BRI", y = "Density") +
  theme_bw(base_size = 13)
```

---

# Train / Test Split

A stratified 70 / 30 split on `low_muscle_mass` is performed to preserve the
~20 % positive-class prevalence in both partitions.  Only complete cases for
the outcome are retained at this stage.

```{r model-variables}
# ── Define modelling feature set ──────────────────────────────────────────────
MODEL_VARS <- c("age_years", "sex", "race_ethnicity",
                "bmi", "weight_kg", "height_cm", "waist_cm",
                "max_grip_kg", "poverty_ratio",
                "protein_g", "energy_kcal", "calcium_mg", "vitamin_d_mcg",
                "bri", "age_group",
                "low_muscle_mass")

nhanes_model <- nhanes_fe %>%
  select(participant_id, all_of(MODEL_VARS)) %>%
  filter(!is.na(low_muscle_mass)) %>%          # outcome must be observed
  mutate(
    low_muscle_mass = factor(low_muscle_mass,
                             levels = c(0, 1),
                             labels = c("Normal", "LowMuscle")),
    sex            = factor(sex),
    race_ethnicity = factor(race_ethnicity),
    age_group      = factor(age_group, levels = c("20-29","30-39","40-49","50-59"))
  ) %>%
  drop_na()    # remove rows with any remaining NA in modelling columns

cat("Complete-case modelling rows:", nrow(nhanes_model), "\n")
cat("Outcome distribution:\n")
print(table(nhanes_model$low_muscle_mass))
```

```{r train-test-split}
# ── Stratified 70/30 split ────────────────────────────────────────────────────
set.seed(42)
train_idx <- createDataPartition(nhanes_model$low_muscle_mass,
                                  p = 0.70, list = FALSE)

train_df <- nhanes_model[ train_idx, ] %>% select(-participant_id)
test_df  <- nhanes_model[-train_idx, ] %>% select(-participant_id)

cat("Training set:", nrow(train_df), "rows\n")
cat("Test set:    ", nrow(test_df),  "rows\n")
cat("\nTraining outcome prevalence:\n")
print(prop.table(table(train_df$low_muscle_mass)))
cat("\nTest outcome prevalence:\n")
print(prop.table(table(test_df$low_muscle_mass)))
```

The stratified split ensures that the positive-class (low muscle mass) ratio is
approximately equal in both partitions, which is important given the ~20 %
prevalence.

---

# Model Training

All four classifiers share the same **5-fold stratified cross-validation**
control object and are trained on identical feature sets.  Class imbalance is
addressed explicitly in each model (class weights or scale parameters).

```{r cv-control}
# ── Shared cross-validation control ──────────────────────────────────────────
set.seed(42)
cv_ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  sampling        = NULL,      # class-weighting handled inside each model
  verboseIter     = FALSE
)

# Class weight ratio (negative : positive) for models that accept it
class_ratio <- sum(train_df$low_muscle_mass == "Normal") /
               sum(train_df$low_muscle_mass == "LowMuscle")
cat(sprintf("Class weight ratio (Normal:LowMuscle) = %.2f\n", class_ratio))

# Helper: model formula (outcome ~ all features)
f_formula <- low_muscle_mass ~ .
```

## Elastic Net Logistic Regression (Baseline)

Elastic net combines L1 (lasso) and L2 (ridge) penalties, enabling automatic
feature selection while handling correlated predictors.  The `alpha` and
`lambda` grid is searched via cross-validation; class imbalance is addressed
through `weights` passed to `glmnet`.

```{r enet-train, cache=TRUE}
# ── Prepare design matrix (caret handles dummies internally for glmnet) ──────
# Build class weights: inverse frequency weighting
wts_train <- ifelse(train_df$low_muscle_mass == "LowMuscle",
                    class_ratio, 1)

set.seed(42)
enet_model <- train(
  f_formula,
  data      = train_df,
  method    = "glmnet",
  trControl = cv_ctrl,
  tuneGrid  = expand.grid(alpha  = c(0, 0.25, 0.5, 0.75, 1),
                           lambda = 10^seq(-4, 0, length.out = 20)),
  weights   = wts_train,
  metric    = "ROC",
  preProcess = c("center", "scale")
)

cat("Best elastic net parameters:\n")
print(enet_model$bestTune)
```

## Random Forest

Random Forest (Breiman, 2001) is an ensemble of 500 decorrelated decision trees.
`mtry` — the number of randomly sampled predictors at each split — is tuned
over a grid; class imbalance is addressed via the `classwt` argument.

```{r rf-train, cache=TRUE}
# Determine number of predictors for mtry grid
n_features <- ncol(train_df) - 1   # subtract outcome

set.seed(42)
rf_model <- train(
  f_formula,
  data      = train_df,
  method    = "rf",
  trControl = cv_ctrl,
  tuneGrid  = data.frame(mtry = c(2, 4, 6, 8, round(sqrt(n_features)),
                                   round(n_features / 3))),
  ntree     = 500,
  classwt   = c("Normal" = 1, "LowMuscle" = class_ratio),
  metric    = "ROC",
  importance = TRUE
)

cat("Best RF mtry:", rf_model$bestTune$mtry, "\n")
```

## XGBoost

XGBoost (Chen & Guestrin, 2016) is a gradient-boosted tree ensemble that
handles class imbalance via `scale_pos_weight` (set to the negative-to-positive
class ratio).  Early stopping is applied within each cross-validation fold to
prevent overfitting.

```{r xgb-train, cache=TRUE}
# ── Encode factors as numeric for xgboost ────────────────────────────────────
encode_df <- function(df) {
  df %>%
    mutate(
      sex_num   = as.integer(sex) - 1L,       # Male=0, Female=1
      race_num  = as.integer(race_ethnicity) - 1L,
      ag_num    = as.integer(age_group) - 1L
    ) %>%
    select(-sex, -race_ethnicity, -age_group)
}

train_xgb <- encode_df(train_df)
test_xgb  <- encode_df(test_df)

# Outcome must be 0/1
y_train <- as.integer(train_xgb$low_muscle_mass == "LowMuscle")
y_test  <- as.integer(test_xgb$low_muscle_mass  == "LowMuscle")

X_train <- as.matrix(train_xgb %>% select(-low_muscle_mass))
X_test  <- as.matrix(test_xgb  %>% select(-low_muscle_mass))

dtrain <- xgb.DMatrix(X_train, label = y_train)
dtest  <- xgb.DMatrix(X_test,  label = y_test)

xgb_params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  eta              = 0.05,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  min_child_weight = 5,
  scale_pos_weight = class_ratio,   # handles class imbalance
  seed             = 42
)

set.seed(42)
xgb_cv <- xgb.cv(
  params   = xgb_params,
  data     = dtrain,
  nrounds  = 500,
  nfold    = 5,
  stratified    = TRUE,
  early_stopping_rounds = 30,
  verbose  = FALSE
)

# Guard: best_iteration is only set when early stopping actually triggered.
# If it is NULL or zero, fall back to the iteration with the best mean test AUC.
best_nrounds <- xgb_cv$best_iteration

if (is.null(best_nrounds) || length(best_nrounds) == 0 || best_nrounds == 0) {
  eval_log    <- xgb_cv$evaluation_log
  best_nrounds <- eval_log$iter[which.max(eval_log$test_auc_mean)]
}

# Final safety net — use 100 rounds if still unresolved
if (is.null(best_nrounds) || length(best_nrounds) == 0 || best_nrounds == 0) {
  best_nrounds <- 100
  warning("Could not determine best_nrounds from CV; defaulting to 100.")
}

cat("XGBoost best number of rounds:", best_nrounds, "\n")

# Train final model on full training data
set.seed(42)
xgb_model <- xgb.train(
  params  = xgb_params,
  data    = dtrain,
  nrounds = best_nrounds,
  verbose = 0
)
```

## Support Vector Machine (Radial Kernel)

An SVM with a radial basis function kernel is tuned over a grid of cost
parameter `C` and kernel width `sigma`.  Class weights are set via `class.weights`.

```{r svm-train, cache=TRUE}
set.seed(42)
svm_model <- train(
  f_formula,
  data      = train_df,
  method    = "svmRadial",
  trControl = cv_ctrl,
  tuneGrid  = expand.grid(C     = c(0.1, 0.5, 1, 2, 5),
                           sigma = c(0.01, 0.05, 0.1, 0.5)),
  class.weights = c("Normal" = 1, "LowMuscle" = class_ratio),
  metric    = "ROC",
  preProcess = c("center", "scale")
)

cat("Best SVM parameters:\n")
print(svm_model$bestTune)
```

---

# Model Evaluation on the Test Set

All models are evaluated on the 30 % held-out test partition.  We report six
metrics that together capture both discrimination and class-specific performance:

| Metric | Rationale |
|--------|-----------|
| AUC | Overall discrimination; threshold-free |
| F1 (macro) | Harmonic mean of precision and recall, averaged over classes |
| MCC | Balanced quality metric; robust to class imbalance |
| Sensitivity | True-positive rate (clinically important) |
| Specificity | True-negative rate |
| F2 | F-measure weighting recall 2× more than precision |

```{r predict-helpers}
# ── Helper: generate predictions and probabilities ────────────────────────────
get_preds <- function(model, newdata, positive = "LowMuscle") {
  probs <- predict(model, newdata = newdata, type = "prob")[, positive]
  preds <- predict(model, newdata = newdata)
  list(prob = probs, class = preds)
}

# ── XGBoost predictions ───────────────────────────────────────────────────────
# Rebuild dtest fresh here: xgb.DMatrix objects cannot survive cache serialisation
dtest_fresh <- xgb.DMatrix(X_test, label = y_test)
xgb_prob    <- predict(xgb_model, dtest_fresh)
xgb_class   <- factor(ifelse(xgb_prob > 0.5, "LowMuscle", "Normal"),
                      levels = c("Normal", "LowMuscle"))

# ── Caret model predictions ───────────────────────────────────────────────────
enet_p <- get_preds(enet_model, test_df)
rf_p   <- get_preds(rf_model,   test_df)
svm_p  <- get_preds(svm_model,  test_df)

truth    <- test_df$low_muscle_mass   # factor with levels Normal / LowMuscle
truth_01 <- as.integer(truth == "LowMuscle")
```

```{r eval-functions}
# ── Metric computation helper ─────────────────────────────────────────────────
compute_metrics <- function(prob, pred_class, truth_factor, truth_01, model_name) {
  cm  <- confusionMatrix(pred_class, truth_factor, positive = "LowMuscle")
  roc_obj <- roc(truth_01, prob, quiet = TRUE)

  sens <- cm$byClass["Sensitivity"]
  spec <- cm$byClass["Specificity"]
  prec <- cm$byClass["Precision"]
  rec  <- cm$byClass["Recall"]

  # F1 macro (average of class-specific F1s)
  f1_pos <- ifelse(!is.na(prec) & !is.na(rec) & (prec + rec) > 0,
                   2 * prec * rec / (prec + rec), NA)
  prec_neg <- cm$byClass["Neg Pred Value"]
  rec_neg  <- spec
  f1_neg   <- ifelse(!is.na(prec_neg) & !is.na(rec_neg) & (prec_neg + rec_neg) > 0,
                     2 * prec_neg * rec_neg / (prec_neg + rec_neg), NA)
  f1_macro <- mean(c(f1_pos, f1_neg), na.rm = TRUE)

  # F2: weights recall twice as heavily as precision
  beta <- 2
  f2   <- ifelse(!is.na(prec) & !is.na(rec) & (beta^2 * prec + rec) > 0,
                 (1 + beta^2) * prec * rec / (beta^2 * prec + rec), NA)

  # Matthews Correlation Coefficient
  # Replace any NA cells (class never predicted) with 0
  tp <- as.numeric(cm$table["LowMuscle", "LowMuscle"]); tp <- ifelse(is.na(tp), 0, tp)
  tn <- as.numeric(cm$table["Normal",    "Normal"]);    tn <- ifelse(is.na(tn), 0, tn)
  fp <- as.numeric(cm$table["LowMuscle", "Normal"]);    fp <- ifelse(is.na(fp), 0, fp)
  fn <- as.numeric(cm$table["Normal",    "LowMuscle"]); fn <- ifelse(is.na(fn), 0, fn)
  mcc_num  <- (tp * tn) - (fp * fn)
  mcc_den  <- sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
  mcc      <- if (is.na(mcc_den) || mcc_den == 0) 0 else mcc_num / mcc_den

  data.frame(
    Model       = model_name,
    AUC         = round(as.numeric(auc(roc_obj)), 4),
    F1_Macro    = round(f1_macro, 4),
    MCC         = round(mcc, 4),
    Sensitivity = round(sens, 4),
    Specificity = round(spec, 4),
    F2          = round(f2, 4)
  )
}
```

```{r compute-all-metrics}
results_tbl <- bind_rows(
  compute_metrics(enet_p$prob, enet_p$class, truth, truth_01, "Elastic Net LR"),
  compute_metrics(rf_p$prob,   rf_p$class,   truth, truth_01, "Random Forest"),
  compute_metrics(xgb_prob,    xgb_class,    truth, truth_01, "XGBoost"),
  compute_metrics(svm_p$prob,  svm_p$class,  truth, truth_01, "SVM (Radial)")
)

# Identify best model by AUC
best_model_name <- results_tbl$Model[which.max(results_tbl$AUC)]
cat("Best model (by AUC):", best_model_name, "\n")
```

```{r results-table}
# ── Formatted results table ───────────────────────────────────────────────────
results_tbl %>%
  gt() %>%
  tab_header(
    title    = "Test-Set Performance of Four Classifiers",
    subtitle = "NHANES 2011–2018 | Outcome: Low Muscle Mass (bottom 20th %ile ASMI)"
  ) %>%
  fmt_number(columns = where(is.numeric), decimals = 4) %>%
  tab_style(
    style = list(cell_fill(color = "#DDEEFF"),
                 cell_text(weight = "bold")),
    locations = cells_body(rows = Model == best_model_name)
  ) %>%
  cols_label(
    Model       = "Model",
    AUC         = "AUC",
    F1_Macro    = "F1 (Macro)",
    MCC         = "MCC",
    Sensitivity = "Sensitivity",
    Specificity = "Specificity",
    F2          = "F2-Score"
  ) %>%
  tab_footnote("Best model (highest AUC) highlighted in blue.")
```

## ROC Curves

```{r roc-curves, fig.cap="ROC curves for all four classifiers on the held-out test set"}
# Build ROC objects
roc_enet <- roc(truth_01, enet_p$prob, quiet = TRUE)
roc_rf   <- roc(truth_01, rf_p$prob,   quiet = TRUE)
roc_xgb  <- roc(truth_01, xgb_prob,    quiet = TRUE)
roc_svm  <- roc(truth_01, svm_p$prob,  quiet = TRUE)

# Combine into a single data frame for ggplot
roc_df <- bind_rows(
  data.frame(FPR = 1 - roc_enet$specificities,
             TPR = roc_enet$sensitivities,
             Model = sprintf("Elastic Net (AUC=%.3f)", auc(roc_enet))),
  data.frame(FPR = 1 - roc_rf$specificities,
             TPR = roc_rf$sensitivities,
             Model = sprintf("Random Forest (AUC=%.3f)", auc(roc_rf))),
  data.frame(FPR = 1 - roc_xgb$specificities,
             TPR = roc_xgb$sensitivities,
             Model = sprintf("XGBoost (AUC=%.3f)", auc(roc_xgb))),
  data.frame(FPR = 1 - roc_svm$specificities,
             TPR = roc_svm$sensitivities,
             Model = sprintf("SVM Radial (AUC=%.3f)", auc(roc_svm)))
)

ggplot(roc_df, aes(x = FPR, y = TPR, colour = Model)) +
  geom_line(linewidth = 1.0) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_brewer(palette = "Set1") +
  labs(title    = "ROC Curves — Test Set Performance",
       subtitle = "Predicting Low Muscle Mass (NHANES 2011–2018)",
       x = "False Positive Rate (1 – Specificity)",
       y = "True Positive Rate (Sensitivity)",
       colour = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = c(0.65, 0.25),
        legend.background = element_rect(fill = "white", colour = "grey80"))
```

---

# SHAP Analysis

SHAP (SHapley Additive exPlanations; Lundberg & Lee, 2017) values provide a
unified measure of each feature's contribution to individual predictions.  We
apply SHAP to the **best-performing model** identified above.

> **Note:** SHAP values are computed using the `kernelshap` package for models
> that do not have native SHAP support (elastic net, SVM); the `shapviz`
> package provides the visualisation layer.  If the best model is XGBoost,
> native tree SHAP values are used instead for exact computation.

```{r shap-setup}
# ── Identify best model object ────────────────────────────────────────────────
best_model_obj <- switch(best_model_name,
  "Elastic Net LR" = enet_model,
  "Random Forest"  = rf_model,
  "SVM (Radial)"   = svm_model,
  "XGBoost"        = xgb_model
)

# ── Two representations of the feature data for SHAP ─────────────────────────
#
# XGBoost path : fully-numeric matrix (factors integer-encoded)
# caret path   : original data frame with factor columns intact,
#                because caret predict() needs them for model.frame()

# 1. Numeric matrix — XGBoost only
shap_prep_numeric <- function(df) {
  df %>%
    select(-low_muscle_mass) %>%
    mutate(
      sex_num  = as.integer(factor(sex, levels = c("Male", "Female"))) - 1L,
      race_num = as.integer(race_ethnicity) - 1L,
      ag_num   = as.integer(age_group) - 1L
    ) %>%
    select(-sex, -race_ethnicity, -age_group) %>%
    as.matrix()
}

test_shap_mat  <- shap_prep_numeric(test_df)
train_shap_mat <- shap_prep_numeric(train_df)
feature_names  <- colnames(test_shap_mat)

# 2. Factor-preserving data frame — caret / kernelshap path
test_shap_df  <- test_df  %>% select(-low_muscle_mass)
train_shap_df <- train_df %>% select(-low_muscle_mass)

cat("SHAP matrix dimensions:    ", dim(test_shap_mat), "\n")
cat("SHAP data frame dimensions:", dim(test_shap_df),  "\n")
```

```{r shap-compute, cache=TRUE}
# ── Prediction wrapper ────────────────────────────────────────────────────────
if (best_model_name == "XGBoost") {

  # Native tree SHAP contributions from XGBoost (exact, fast)
  # Rebuild DMatrix fresh — cached DMatrix objects are disposed between chunks
  dtest_shap <- xgb.DMatrix(test_shap_mat)
  shap_vals  <- predict(xgb_model,
                        dtest_shap,
                        predcontrib   = TRUE,
                        approxcontrib = FALSE)
  # Remove the bias column (last column named "BIAS")
  shap_vals <- shap_vals[, seq_len(ncol(shap_vals) - 1)]
  colnames(shap_vals) <- feature_names

} else {

  # kernelshap for caret models (maintained successor to fastshap)
  # kernelshap passes rows of X / bg_X directly to pred_fun.
  # We use the factor-preserving data frames so caret's model.frame()
  # can find the original columns (sex, race_ethnicity, age_group).
  pred_wrapper <- function(object, newdata) {
    # newdata arrives as a data frame slice — ensure factors are correct
    nd <- as.data.frame(newdata)
    nd$sex            <- factor(nd$sex,
                                levels = levels(train_df$sex))
    nd$race_ethnicity <- factor(nd$race_ethnicity,
                                levels = levels(train_df$race_ethnicity))
    nd$age_group      <- factor(nd$age_group,
                                levels = levels(train_df$age_group))
    predict(object, newdata = nd, type = "prob")[, "LowMuscle"]
  }

  # Use a 100-row background sample from the factor-preserving training frame
  set.seed(42)
  bg_idx <- sample(nrow(train_shap_df), min(100, nrow(train_shap_df)))

  set.seed(42)
  ks_obj <- kernelshap(
    object   = best_model_obj,
    X        = test_shap_df,          # factor data frame
    bg_X     = train_shap_df[bg_idx, ],
    pred_fun = pred_wrapper,
    verbose  = FALSE
  )
  shap_vals <- ks_obj$S   # matrix of SHAP values (n_test x n_features)
  colnames(shap_vals) <- colnames(test_shap_df)
}

# ── Build shapviz object ──────────────────────────────────────────────────────
# XGBoost path uses the numeric matrix; caret path uses the factor data frame.
# shapviz only needs X for display labels, so either works.
if (best_model_name == "XGBoost") {
  sv <- shapviz(shap_vals, X = test_shap_mat)
} else {
  sv <- shapviz(shap_vals, X = as.data.frame(test_shap_df))
}
cat("SHAP values computed for", nrow(shap_vals), "test observations.\n")
```

## Global SHAP Beeswarm Summary

```{r shap-beeswarm, fig.cap="Global SHAP beeswarm plot: each point is one test observation; colour encodes feature value; width shows distribution of impact."}
sv_importance(sv, kind = "beeswarm", max_display = 15L) +
  labs(title    = paste("SHAP Beeswarm Summary —", best_model_name),
       subtitle = "Each point = one observation; colour = feature value (red=high, blue=low)") +
  theme_bw(base_size = 12)
```

## Mean |SHAP| Bar Plot

```{r shap-bar, fig.cap="Mean absolute SHAP value per feature — indicates average contribution to model output."}
sv_importance(sv, kind = "bar", max_display = 15L) +
  labs(title    = paste("Mean |SHAP| Values —", best_model_name),
       subtitle = "Average absolute contribution to predicted log-odds of Low Muscle Mass") +
  theme_bw(base_size = 12)
```

```{r shap-identify-cases}
# ── Identify one TP, one FN, one FP from test set ────────────────────────────
# We need predicted class and true class for the best model

if (best_model_name == "XGBoost") {
  test_pred_class <- xgb_class
  test_pred_prob  <- xgb_prob
} else {
  preds_obj       <- get_preds(best_model_obj, test_df)
  test_pred_class <- preds_obj$class
  test_pred_prob  <- preds_obj$prob
}

outcomes_df <- data.frame(
  idx        = seq_len(nrow(test_df)),
  truth      = truth,
  pred       = test_pred_class,
  prob       = test_pred_prob
)

tp_idx <- outcomes_df %>% filter(truth == "LowMuscle", pred == "LowMuscle") %>%
          slice(1) %>% pull(idx)
fn_idx <- outcomes_df %>% filter(truth == "LowMuscle", pred == "Normal") %>%
          slice(1) %>% pull(idx)
fp_idx <- outcomes_df %>% filter(truth == "Normal",    pred == "LowMuscle") %>%
          slice(1) %>% pull(idx)

cat("True Positive  index:", tp_idx, "\n")
cat("False Negative index:", fn_idx, "\n")
cat("False Positive index:", fp_idx, "\n")
```

## SHAP Waterfall Plots

```{r shap-waterfall-tp, fig.cap="SHAP waterfall for a True Positive: model correctly predicts Low Muscle Mass."}
sv_waterfall(sv, row_id = tp_idx) +
  labs(title = paste0("True Positive (obs #", tp_idx, ") — Predicted: LowMuscle | Actual: LowMuscle")) +
  theme_bw(base_size = 12)
```

```{r shap-waterfall-fn, fig.cap="SHAP waterfall for a False Negative: model incorrectly predicts Normal for a Low Muscle Mass case."}
sv_waterfall(sv, row_id = fn_idx) +
  labs(title = paste0("False Negative (obs #", fn_idx, ") — Predicted: Normal | Actual: LowMuscle")) +
  theme_bw(base_size = 12)
```

```{r shap-waterfall-fp, fig.cap="SHAP waterfall for a False Positive: model incorrectly predicts Low Muscle Mass for a Normal case."}
sv_waterfall(sv, row_id = fp_idx) +
  labs(title = paste0("False Positive (obs #", fp_idx, ") — Predicted: LowMuscle | Actual: Normal")) +
  theme_bw(base_size = 12)
```

The waterfall plots decompose each prediction into feature-level additive
contributions from the model's base value (average log-odds across training
data).  Red bars push the prediction toward LowMuscle; blue bars push it toward
Normal.

---

# Subgroup Analysis

We stratify test-set AUC, F1 (macro), and MCC by three grouping variables:
`sex`, `age_group`, and `race_ethnicity`.  This reveals whether model
performance is equitable across population subgroups.

```{r subgroup-function}
# ── Generalised subgroup metric function ──────────────────────────────────────
subgroup_metrics <- function(group_var, test_data, prob_vec, class_vec) {
  groups <- unique(test_data[[group_var]])
  groups <- sort(as.character(groups))

  purrr::map_dfr(groups, function(g) {
    idx   <- which(as.character(test_data[[group_var]]) == g)
    if (length(idx) < 10) return(NULL)   # skip tiny strata

    sub_truth01 <- as.integer(test_data$low_muscle_mass[idx] == "LowMuscle")
    sub_prob    <- prob_vec[idx]
    sub_pred    <- class_vec[idx]
    sub_truth   <- test_data$low_muscle_mass[idx]

    if (length(unique(sub_truth01)) < 2) return(NULL)  # need both classes

    roc_sub <- roc(sub_truth01, sub_prob, quiet = TRUE)
    cm_sub  <- confusionMatrix(sub_pred, sub_truth, positive = "LowMuscle")

    sens <- cm_sub$byClass["Sensitivity"]
    spec <- cm_sub$byClass["Specificity"]
    prec <- cm_sub$byClass["Precision"]
    rec  <- cm_sub$byClass["Recall"]
    f1_pos <- ifelse(!is.na(prec) & !is.na(rec) & (prec + rec) > 0,
                     2 * prec * rec / (prec + rec), NA)
    prec_neg <- cm_sub$byClass["Neg Pred Value"]
    rec_neg  <- spec
    f1_neg   <- ifelse(!is.na(prec_neg) & !is.na(rec_neg) & (prec_neg + rec_neg) > 0,
                       2 * prec_neg * rec_neg / (prec_neg + rec_neg), NA)
    f1_macro <- mean(c(f1_pos, f1_neg), na.rm = TRUE)

    tp <- as.numeric(cm_sub$table["LowMuscle", "LowMuscle"]); tp <- ifelse(is.na(tp), 0, tp)
    tn <- as.numeric(cm_sub$table["Normal",    "Normal"]);    tn <- ifelse(is.na(tn), 0, tn)
    fp <- as.numeric(cm_sub$table["LowMuscle", "Normal"]);    fp <- ifelse(is.na(fp), 0, fp)
    fn <- as.numeric(cm_sub$table["Normal",    "LowMuscle"]); fn <- ifelse(is.na(fn), 0, fn)
    mcc_den <- sqrt((tp + fp) * (tp + fn) * (tn + fp) * (tn + fn))
    mcc <- if (is.na(mcc_den) || mcc_den == 0) 0 else ((tp * tn) - (fp * fn)) / mcc_den

    data.frame(
      Grouping_Variable = group_var,
      Subgroup          = g,
      N                 = length(idx),
      Prevalence        = round(mean(sub_truth01) * 100, 1),
      AUC               = round(as.numeric(auc(roc_sub)), 4),
      F1_Macro          = round(f1_macro, 4),
      MCC               = round(mcc, 4)
    )
  })
}
```

```{r subgroup-compute, message=FALSE}
# Add age_group to test_df for subgroup analysis
test_df_sg <- test_df %>%
  mutate(age_group = cut(age_years,
                         breaks = c(20, 30, 40, 50, 60),
                         labels = c("20-29","30-39","40-49","50-59"),
                         right  = FALSE, include.lowest = TRUE))

# Use best model predictions
sg_sex  <- subgroup_metrics("sex",            test_df_sg, test_pred_prob, test_pred_class)
sg_age  <- subgroup_metrics("age_group",      test_df_sg, test_pred_prob, test_pred_class)
sg_race <- subgroup_metrics("race_ethnicity", test_df_sg, test_pred_prob, test_pred_class)

subgroup_results <- bind_rows(sg_sex, sg_age, sg_race)
```

```{r subgroup-table}
subgroup_results %>%
  gt(groupname_col = "Grouping_Variable") %>%
  tab_header(
    title    = "Subgroup Analysis of Best Model Performance",
    subtitle = paste("Model:", best_model_name,
                     "| Metric: AUC, F1 (Macro), MCC")
  ) %>%
  fmt_number(columns = c(AUC, F1_Macro, MCC), decimals = 4) %>%
  cols_label(
    Subgroup   = "Subgroup",
    N          = "N (test)",
    Prevalence = "LMM Prevalence (%)",
    AUC        = "AUC",
    F1_Macro   = "F1 (Macro)",
    MCC        = "MCC"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_footnote("LMM = Low Muscle Mass. Subgroups with < 10 observations are excluded.")
```

```{r subgroup-plot, fig.cap="AUC by subgroup across sex, age group, and race/ethnicity"}
subgroup_results %>%
  mutate(label = paste0(Subgroup, "\n(n=", N, ")"),
         Grouping_Variable = factor(Grouping_Variable,
                                    levels = c("sex", "age_group", "race_ethnicity"),
                                    labels = c("Sex", "Age Group", "Race/Ethnicity"))) %>%
  ggplot(aes(x = reorder(label, AUC), y = AUC, fill = Grouping_Variable)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = max(results_tbl$AUC[results_tbl$Model == best_model_name]),
             linetype = "dashed", colour = "grey40") +
  coord_flip() +
  facet_wrap(~ Grouping_Variable, scales = "free_y", ncol = 1) +
  scale_fill_brewer(palette = "Set2") +
  labs(title    = "Subgroup AUC — Fairness Assessment",
       subtitle = paste("Best model:", best_model_name),
       x = NULL, y = "AUC",
       fill = "Grouping Variable") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

# Interaction Term Analysis

We extend the elastic net logistic regression baseline with **pairwise
interaction terms** between the three grouping variables (`sex`, `age_group`,
`race_ethnicity`) and the **top 5 SHAP predictors** identified in Section 7.
Statistical significance is assessed using **likelihood ratio tests (LRT)**
comparing nested models; we report chi-square statistics and p-values.

```{r top5-shap}
# ── Extract top 5 predictors by mean |SHAP| ───────────────────────────────────
mean_shap <- colMeans(abs(shap_vals))
top5_shap <- names(sort(mean_shap, decreasing = TRUE))[1:5]
cat("Top 5 SHAP predictors:\n")
print(top5_shap)
```

```{r interaction-data}
# ── Prepare data for logistic regression (with dummies) ──────────────────────
# Use complete-case training data; glm handles factors natively

# Prepare a version of training data suitable for glm
train_glm <- train_df %>%
  mutate(
    outcome_01     = as.integer(low_muscle_mass == "LowMuscle"),
    sex            = factor(sex),
    race_ethnicity = factor(race_ethnicity),
    age_group      = factor(age_group,
                            levels = c("20-29","30-39","40-49","50-59"))
  )

test_glm <- test_df %>%
  mutate(
    outcome_01     = as.integer(low_muscle_mass == "LowMuscle"),
    sex            = factor(sex),
    race_ethnicity = factor(race_ethnicity),
    age_group      = factor(age_group,
                            levels = c("20-29","30-39","40-49","50-59"))
  )

# Map SHAP feature names back to original column names
# (SHAP matrix may use encoded names; map them back)
shap_to_orig <- c(
  sex_num  = "sex",
  race_num = "race_ethnicity",
  ag_num   = "age_group"
)

top5_orig <- dplyr::recode(top5_shap, !!!shap_to_orig)
cat("Top 5 (original names):", top5_orig, "\n")
```

```{r base-logistic}
# ── Base logistic model (no interactions) ────────────────────────────────────
# Use the top 5 SHAP predictors plus grouping variables
base_vars  <- unique(c(top5_orig, "sex", "age_group", "race_ethnicity"))

# Build formula safely — keep only vars present in data
base_vars  <- base_vars[base_vars %in% names(train_glm)]

base_f     <- as.formula(paste("outcome_01 ~",
                               paste(base_vars, collapse = " + ")))

m_base <- glm(base_f, data = train_glm, family = binomial,
              weights = ifelse(train_glm$outcome_01 == 1, class_ratio, 1))

cat("Base model AIC:", round(AIC(m_base), 1), "\n")
cat("Base model df:", df.residual(m_base), "\n")
```

```{r interaction-lrt}
# ── Add interaction terms and run LRT ────────────────────────────────────────
group_vars <- c("sex", "age_group", "race_ethnicity")

# Keep only top5 predictors that are NOT already a grouping variable
top5_non_group <- top5_orig[!top5_orig %in% group_vars]

lrt_results <- list()

for (gv in group_vars) {
  for (fv in top5_non_group) {

    # Skip if feature not in data
    if (!fv %in% names(train_glm)) next

    # Extended model formula: add the interaction term
    int_term <- paste0(gv, ":", fv)
    ext_f    <- update(base_f,
                       paste(". ~ . +", int_term))

    tryCatch({
      m_ext   <- glm(ext_f, data = train_glm, family = binomial,
                     weights = ifelse(train_glm$outcome_01 == 1, class_ratio, 1))
      lrt_out <- anova(m_base, m_ext, test = "LRT")

      lrt_results[[paste(gv, fv, sep = "_x_")]] <- data.frame(
        Grouping_Var = gv,
        Predictor    = fv,
        Interaction  = paste0(gv, " × ", fv),
        Chi_Square   = round(lrt_out$Deviance[2], 4),
        df           = lrt_out$Df[2],
        p_value      = round(lrt_out[["Pr(>Chi)"]][2], 6)
      )
    }, error = function(e) {
      message("LRT failed for ", gv, " × ", fv, ": ", e$message)
    })
  }
}

lrt_tbl <- bind_rows(lrt_results) %>%
  arrange(p_value) %>%
  mutate(
    Significant   = if_else(p_value < 0.05, "Yes", "No"),
    p_value_label = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01  ~ "< 0.01",
      p_value < 0.05  ~ "< 0.05",
      TRUE            ~ as.character(round(p_value, 4))
    )
  )

cat("Likelihood ratio test results:\n")
print(lrt_tbl %>% select(Interaction, Chi_Square, df, p_value_label, Significant))
```

```{r lrt-table}
lrt_tbl %>%
  select(Interaction, Chi_Square, df, p_value_label, Significant) %>%
  gt() %>%
  tab_header(
    title    = "Likelihood Ratio Tests for Interaction Terms",
    subtitle = "Grouping variables (sex, age group, race/ethnicity) × Top 5 SHAP predictors"
  ) %>%
  cols_label(
    Interaction   = "Interaction Term",
    Chi_Square    = "χ² Statistic",
    df            = "df",
    p_value_label = "p-value",
    Significant   = "p < 0.05?"
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#DDEEFF"),
                 cell_text(weight = "bold")),
    locations = cells_body(rows = Significant == "Yes")
  ) %>%
  tab_footnote(
    "χ² and p-values from likelihood ratio tests comparing base model to model
    with added interaction term. Significant terms (p < 0.05) highlighted in blue.
    Bonferroni-corrected threshold: p < 0.05 / number of tests."
  )
```

```{r bonferroni-note}
n_tests    <- nrow(lrt_tbl)
bonf_alpha <- 0.05 / n_tests
cat(sprintf("Bonferroni-corrected significance threshold (alpha / %d tests): %.5f\n",
            n_tests, bonf_alpha))

bonf_sig <- lrt_tbl %>% filter(p_value < bonf_alpha)
cat("Interaction terms significant after Bonferroni correction:\n")
if (nrow(bonf_sig) > 0) {
  print(bonf_sig %>% select(Interaction, Chi_Square, p_value_label))
} else {
  cat("  None survived Bonferroni correction.\n")
}
```

Interaction terms that survive Bonferroni correction indicate that the effect
of a predictor on low-muscle-mass risk meaningfully varies by sex, age decade,
or racial/ethnic group — a clinically important finding for targeted
interventions.

---

# Session Information

```{r session-info}
sessionInfo()
```

---

*Analysis conducted for CIND860 Capstone Project, Toronto Metropolitan University.*
*Data source: CDC NHANES 2011–2018. Adiel Mallik, `r format(Sys.Date(), "%B %Y")`.*
